The work conducted during this thesis has led to scientific contributions disseminated through peer-reviewed international journal and conferences.

\subsection*{International Journals}

\begin{itemize}
    \item \textbf{J. R. Ky}, B. Mathieu, A. Lahmadi, \& R. Boutaba. "ML models for detecting QoE degradation in low-latency applications: a cloud-gaming case study". \textit{IEEE Transactions on Network and Service Management}.
    \item X. Marchal, P. Graff, \textbf{J. R. Ky}, T. Cholez, S. Tuffin, B. Mathieu, \& O. Festor. An analysis of cloud gaming platforms behaviour under synthetic network constraints and real cellular networks conditions. Journal of Network and Systems Management, 31(2), 39.
\end{itemize}


\subsection*{International Conferences}

\joel{TODO}

\begin{itemize}
    \item \textbf{J. R. Ky}, B. Mathieu, A. Lahmadi, M. Wang, N. Marrot, \& R. Boutaba. "Root cause diagnosis in Cloud VR applications over Wi-Fi networks". \textit{Currently under review at XXX}
    
    \item \textbf{J. R. Ky}, B. Mathieu, A. Lahmadi, \& R. Boutaba. "CATS: Contrastive learning for Anomaly detection in Time Series", in \textit{IEEE International Conference on Big Data (Big Data), IEEE.}, Washington DC, USA, December 15-18, 2024.
    
    \item B. Mathieu, O. Dugeon, \textbf{J. R. Ky}, P. Graff, \& T. Cholez. "Segment Routing for Chaining Micro-Services at Different Programmable Network Levels", in \textit{27th Conference on Innovation in Clouds, Internet and Networks (ICIN), IEEE}, Paris, France, March 11-14, 2024.

    \item \textbf{J. R. Ky}, P. Graff, B. Mathieu, \& T. Cholez. "A hybrid P4/NFV architecture for cloud gaming traffic detection with unsupervised ML", in \textit{IEEE Symposium on Computers and Communications (ISCC), IEEE}, Gammarth, Tunisia, July 9-12, 2023.

    \item \textbf{J. R. Ky}, B. Mathieu, A. Lahmadi, \& R. Boutaba. "Assessing unsupervised machine learning solutions for anomaly detection in cloud gaming sessions", in \textit{Workshop on High-Precision, Predictable, and Low-Latency Networking (HiPNet ‘22), colocated with 18th International Conference on Network and Service Management (CNSM), IEEE}, Thessaloniki, Greece, October 31 - November 4, 2022. 
\end{itemize}